---
title: Reeeeeeeeeeeegex!!!
author: ''
date: '2018-03-26'
slug: reeeeeeeeeeeegex
categories: []
tags:
  - R
  - regex
---

What is `regex`? The official defintion from [wikipedia](https://en.wikipedia.org/wiki/Regular_expression) is _a sequence of characters that define a search pattern._ In English, it's basically a way of describing the structure of a character string and it's the most difficult thing to remember. So this a short and ever evolving post to store all the important `regex` or regular expressions I come across and expect to use often.

I use `regex` most often when I need to replace some string during some data cleaning. My current favourite way to do this is to use `stringr::str_replace_all`. 

Use `\\` to be able to reference (escape) special `R` characters like `*`, `\` and `.`.

```{r}
library(stringr)

fake_word <- "Hell0, m! n@m* is T0ju"
str_replace(fake_word, "\\*", "X")

```

Or to replace everything after a certain pattern

```{r}
str_replace(fake_word, "m.*", "")
```

## My favourite resources

As always here are some great resources I keep coming back to:

1. [RStudio cheatsheet](https://www.rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf)

I also heard of an RStudio add-in that helps you with this and I thought I bookmarked the tweet but can't find it right now.

And lastly, here's Missy:

<iframe width="560" height="315" src="https://www.youtube.com/embed/Ijn6scQ70xQ" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>